---
title: "Linear modelling of soil temperature effects on root lesion nematode population densities in R"
author:
- name: Adam H Sparks
  affiliation: University of Southern Queensland, Centre for Crop Health, Toowoomba, Qld, AU
- name: John P Thompson
  affiliation: University of Southern Queensland, Centre for Crop Health, Toowoomba, Qld, AU
output: html_document
bibliography: bibliography.bib
csl: phytopathology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[![DOI](https://zenodo.org/badge/119438897.svg)](https://zenodo.org/badge/latestdoi/119438897)

# Introduction

_Pratylenchus thornei_, the root-lesion nematode, is widely distributed in wheat
(_Triticum aestivum_) growing areas of many countries and is of particular
concern in sub-tropical environments [@THOMPSON2015]. These nematodes penetrate
roots to  feed and reproduce in the root cortex leading to loss of root
function, which affects nutrient and water uptake of nutrients and water causing
nutrient deficiency and water stress [@THOMPSON2015].

In the original paper the population response of _P. thornei_ in Queensland,
Australia wheat to temperature is modelled using a linear and quadratic
equations. The study aimed to investigate the effects of soil profile
temperatures after different sowing dates on reproduction of the nematodes in
susceptible and moderately resistant wheat cultivars in the subtropical grain
region of eastern Australia. This document recreates the models for population
densities of _P. thornei_ as described in the original paper.

## Objectives

There are two types of models described in the paper, the first model is a
linear model used to describe the unplanted control and two quadratic models fit
Gatcher (Susceptible) and GS50a (Moderately Resistant) wheat cultivars. For a
more detailed discussion on fitting plant disease models in R, please see the
"[Linear Regression](http://www.apsnet.org/edcenter/advanced/topics/EcologyAndEpidemiologyInR/DiseaseProgress/Pages/LinearRegression.aspx)" module in the "Ecology and Epidemiology
in R" documents available in the American Phytopathological Society's (APS)
Education Center. For an even more in-depth discussion on linear models in R,
how to fit and how to interpret the diagnostics that R provides the reader
should refer to Faraway [-@FARAWAY2002].

This post will illustrate how to fit the original linear and quadratic models
using the original data in R [@R2017].

# Packages

Using the **`tidyverse`**, [-@TIDYVERSE2017] package simplifies the libraries.
It includes, **`readr`** [-@READR2017], used to import the data; **`tidyr`**
[-@TIDYR2018], used to format the data; **`dplyr`** [-@DPLYR2017], used to
subset the data; and **`ggplot2`** [-@GGPLOT22009], used for visualising the
data and models.

```{r libraries_and_seed, message=FALSE}
library(tidyverse)
```

# Data Wrangling

Import the data using `read_csv()` from **`readr`**. The data are located in the
`data` subfolder.

```{r data_import, echo=TRUE, message=FALSE}
nema <- read_csv("data/Degree Days Relationships.csv")

nema
```

### Description of Fields in the Data

* **Weeks** Number of weeks after wheat sowing

* **Days** Number of days after wheat sowing

* **Temperature** Temperature (˚C) treatment

* **Degree_days** Average thermal time degree days above 10 ˚C for four soil
depths (8, 15, 30 and 60 cm)

* **Unplanted** Log nematode population in the control treatment with no wheat
planted

* **Gatcher** Log nematode population in a susceptible wheat cultivar

* **GS50a** Log nematode population moderately resistant wheat cultivar

* **Potam** Log nematode population susceptible wheat cultivar

* **Suneca** Log nematode population susceptible wheat cultivar

### Wide to Long Data

You can see that each of the varieties have their own column in the original
data format (wide). Using `gather()` from the **`tidyr`** package, convert from
to long format where the varieties are all listed in a single column, "Variety".
The `data` parameter tells R which data frame to gather. The `key` parameter is
the name of the new column to be created called "Variety", `value` specifies the
column that will contain the values that go with the varieties, "Log_pop".
The last portion tells `gather()` which columns are to be gathered. Using the
operator `:` means take the columns from "Unplanted" to "Suneca" and gather
them without needing to type all the column names.

```{r gather_data, echo=TRUE}
nema_long <-
  gather(data = nema,
         key = Variety,
         value = Log_pop,
         Unplanted:Suneca)

nema_long
```

# Data Visulalisation

Now that the data are in the format that **`ggplot2`** uses, take a look at
the data first to see what it looks like. Fit a smoothed line for each variety's
log nematode population to the raw data. The individual temperature treatments
are shown here by shape, the variety by colour.

```{r raw_data_scatterplots, echo=TRUE}
ggplot(
  nema_long,
  aes(
    x = Degree_days,
    y = Log_pop,
    colour = Temperature,
    group = Variety
  )
) +
  geom_point() +
  geom_smooth(colour = "grey",
              se = FALSE,
              alpha = 0.5) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_colour_viridis_c("Temperature") +
  facet_wrap( ~ Variety, ncol = 2)
```

# Modelling

## Unplanted Model

The paper uses a linear model for the unplanted control. Here write a function
to use in modelling the unplanted population data. I have wrapped the model
in a function which makes it pipe-able, `%>%` and has other advantages when it
comes to fitting the same model to several sets of data.

In the linear equation for the Unplanted control treatment, the rate of
population increase can be expressed as:

$$y = y_0 + rt$$


Where $y_0$ is the initial population, $r$ is the rate of change and $t$
equal time.

### Fitting a Linear Model

```{r linear_model, echo=TRUE}
linear_model <- function(df) {
  lm(Log_pop ~ Degree_days,
     data = df)
}
```

Now check the model fit, using `filter()` from **`dplyr`** to select only
Unplanted data from the data set for the model and fit the linear model to the
data. Using `par(mfrow = c(2, 2))` sets up a four-panel graph,
`plot(unplanted_model)` will graph the diagnostic plots for your inspection. For
a more detailed look at how to interpret these graphs, please refer to Faraway
(2002).

```{r check_model}
unplanted_model <- nema_long %>%
  filter(Variety == "Unplanted") %>%
  linear_model()

par(mfrow = c(2, 2))
plot(unplanted_model)

summary(unplanted_model)
```

#### Visualising the Model Fit to the Data

Using **`ggplot2`**'s `geom_smooth()` we can fit the same model above and graph
the resulting line and standard error.

```{r visualise_linear}
nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Unplanted") %>%
  ggplot(aes(
    x = Degree_days,
    y = Log_pop,
    colour = Temperature
  )) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x,
              size = 1,
              se = FALSE,
              colour = "grey",
              alpha = 0.5) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_colour_viridis_c("Temperature") +
  ggtitle("Unplanted Linear Model")
```

## Quadratic Models

In the original paper, original paper, the quadratic model best described
Gatcher and GS50a data, which are fit here.

```{r quadratic_models, echo=TRUE}
quadratic_model <- function(df) {
  lm(Log_pop ~ Degree_days + I(Degree_days^2),
      data = df)
}
```

### Susceptible Varieties

Gatcher, Potam and Suneca all have very similar curves, here Gatcher is used to
fit a quadratic model as in the original paper following the same methods as
above for the linear model.

```{r susceptible_model, echo=TRUE}
s_model <- nema_long %>%
  filter(Variety == "Gatcher") %>% 
  quadratic_model()

par(mfrow = c(2, 2))
plot(s_model)

summary(s_model)
```

#### Visualise Susceptible Variety Model

The model visualisation is the same for the quadratic models as the linear
model, however you will note that the line has a downward curve at higher
temperatures.

```{r visualise_s_model}
nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Gatcher") %>%
  ggplot(aes(
    x = Degree_days,
    y = Log_pop,
    colour = Temperature,
  )) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              size = 1,
              se = FALSE,
              colour = "grey",
              alpha = 0.5) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_colour_viridis_c("Temperature") +
  ggtitle("Gatcher Quadratic Model")
```


### Moderately Resistant Cultiver

GS50a, moderately resistant to _P. thornei_, also fits a quadratic model but the
coefficients are slightly different due to different responses to the variety
and temperature.

```{r moderately_resistant_model, echo=TRUE}
mr_model <- nema_long %>%
  filter(Variety == "GS50a") %>%
  quadratic_model()

par(mfrow = c(2, 2))
plot(mr_model)

summary(mr_model)
```

#### Visualising the Model Fit to the Data

```{r visualise_mr_model}
nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "GS50a") %>%
  ggplot(aes(
    x = Degree_days,
    y = Log_pop,
    colour = Temperature,
  )) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              size = 1,
              se = FALSE,
              colour = "grey",
              alpha = 0.5) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_colour_viridis_c("Temperature") +
  ggtitle("GS50a Quadratic Model")
```

# Discussion and Conclusions

As in the original paper, the model equations can be derived from these models
as well. The derived regression equations are:

Gatcher (Susceptible): $$ln(P. thornei + 1) = -0.000003(0.0000009)T^2 + 0.009(0.0019)T + 5.4671(0.904)$$

GS50a (Moderately Resistant): $$ln(P. thornei + 1) = -0.000002(0.0000007)T^2 + 0.0063(0.0014)T + 5.1559(0.678)$$

Unplanted Control: $$ln(P. thornei + 1) = 0.0013(0.00018)T + 5.4151(0.193)$$

See the `summary()` outputs for each of the models for the coefficient values
and $R^2$ values, which match those reported in the original paper where the
models were fit with Genstat.

# Recommended Reading

# Reproducibility

```{r reproducibility, echo=FALSE}
devtools::session_info()
```

# References
