---
title: "Fitting Linear and Quadratic Models of Nematode Population Responses to Time and Temperature"
author:
- name: Adam H Sparks
  affiliation: University of Southern Queensland, Centre for Crop Health, Toowoomba, Qld, AU
- name: John P Thompson
  affiliation: University of Southern Queensland, Centre for Crop Health, Toowoomba, Qld, AU
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Thompson, JP, 2015. Modelling population densities of root-lesion nematode
_Pratylenchus thornei_ from soil profile temperatures to choose an optimum
sowing date for wheat in a subtropical region. _Field Crops Research_ 183:50-55
DOI: 10.1016/j.fcr.2015.07.005. URL:
<http://www.sciencedirect.com/science/article/pii/S0378429015300083>

# Introduction

_Pratylenchus thornei_, the root-lesion nematode is widely distributed in wheat
(_Triticum aestivum_) growing areas of many countries and is of particular
concern in sub-tropical environments (Thompson 2015). These nematodes penetrate
roots to  feed and reproduce in the root cortex leading to loss of root
function, which affects nutrient and water uptake of nutrients and water causing
nutrient deficiency and water stress (Thompson 2015).

In the original paper the population of _P. thornei_ in wheat in Queensland,
Australia is modelled using a linear and quadratic equations. The study aimed to
investigate the effects of soil profile temperatures after different sowing
dates on reproduction of the nematodes in susceptible and moderately resistant
wheat cultivars in the subtropical grain region of eastern Australia. This
document recreates the models for population densities of _P. thornei_ as
described in _Modelling population densities of root-lesion nematode_ (P.
thornei) _from soil profile temperatures to choose an optimum sowing date for
wheat_ (Thompson 2015).

There are two types of models described in the paper, the first model is a
linear model used to describe the unplanted control and two quadratic models fit
Gatcher (Susceptible) and GS50a (Moderately Resistant) wheat cultivars.

## Session Setup

Using the **`tidyverse`** package simplifies the libraries. It includes,
**`readr`**, used to import the data; **`tidyr`**, used to format the data;
**`dplyr`**, used to subset the data; and **`ggplot2`**, used for visualising
the data and models.

```{r libraries_and_seed, message=FALSE}
library(tidyverse)
```

We will use the `set.seed` function for reproducibility.

```
set.seed(52)
```

## Data Import and Inspection

Import the data using `read_csv()` from **`readr`**.

```{r data_import, echo=TRUE, message=FALSE}
nema <- read_csv("data/Degree Days Relationships.csv")

nema
```

### Description of Fields in the Data

* **Weeks** Number of weeks after planting

* **Days** Number of days after planting

* **Temperature** Temperature (˚C) Treatment

* **Degree_days** Degree days above 10 ˚C

* **Unplanted** Log nematode population in the control treatment with no wheat
planted

* **Gatcher** Log nematode population in a susceptible wheat cultivar

* **GS50a** Log nematode population moderately resistant wheat cultivar

* **Potam** Log nematode population susceptible wheat cultivar

* **Suneca** Log nematode population susceptible wheat cultivar

### Wide to Long Data

You can see that each of the varieties have their own column in the original
data format (wide). Using `gather()` from the **`tidyr`** package, convert from
to long format where the varieties are all listed in a single column, "Variety".
The `data` paramter tells R which data frame to gather. The `key` parameter is
the name of the new column to be created called "Variety", `value` specifies the
column that will contain the values that go with the varieties, "Population".
The last portion tells `gather()` which columns are to be gathered. Using the
operator `:` means take the columns from "Unplanted" to "Suneca" and gather
them without needing to type all the column names.

```{r gather_data, echo=TRUE}
nema_long <-
  gather(data = nema,
         key = Variety,
         value = Population,
         Unplanted:Suneca)

nema_long
```

Now that the data are in the format that **`ggplot2`** prefers, take a look at
the data first to see what it looks like. Fit a smoothed line for each variety's
log nematode population to the raw data. The individual temperature treatments
are shown here by shape, the variety by colour.

```{r raw_data_scatterplots, echo=TRUE}
ggplot(
  nema_long,
  aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )
) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth() +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5))
```

# Model Fitting

## Unplanted Model

The paper uses a linear model for the unplanted control. Here write a function
to use in modelling the unplanted population data. I have wrapped the model
in a function which makes it pipeable, `%>%` and has other advantages when it
comes to fitting the same model to several sets of data.

Linear models in R are fit using the `lm()` function with the form of

$$y = a x + b$$

In this case the rate of population growth can be represented as

$$y = y_0 + rt$$


Where $y_0$ is the initial population, $r$ is the rate of change and $t$
equal time.

### Linear Model

```{r linear_model_HT, echo=TRUE}
linear_model <- function(df) {
  lm(Population ~ Degree_days,
     data = df)
}
```

Now check the model fit, using `filter()` from **`dplyr`** to select only
Unplanted data from the data set for the model and fit the linear model to the
data.

Lastly we can show the unplanted data alone as a scatterplot with the model
line fitted using `geom_smooth()`.

```{r check_model}
unplanted_model <- nema_long %>%
  filter(Variety == "Unplanted") %>%
  linear_model()

par(mfrow = c(2, 2))
plot(unplanted_model)

summary(unplanted_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Unplanted") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("Unplanted Linear Model")
```

## Quadratic Models

In the original paper, Gatcher and GS50a best fit quadratic models, which are
fit here.

```{r HT_quadratic_models, echo=TRUE}
quadratic_model <- function(df) {
  lm(Population ~ Degree_days + poly(Degree_days, 2),
      data = df)
}
```

### Susceptible Varieties

Gatcher, Potam and Suneca all have very similar curves, here Gatcher is used to
fit a quadratic model as in the original paper following the same methods as
above for the linear model.

```{r susceptible_model, echo=TRUE}
s_model <- nema_long %>%
  filter(Variety == "Gatcher") %>% 
  quadratic_model()

par(mfrow = c(2, 2))
plot(s_model)

summary(s_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Gatcher") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x + poly(x, 2),
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("Gatcher Quadratic Model")
```

### Moderately Resistant Cultiver

GS50a, moderately resistant to _P. thornei_, also fits a quadratic model but the
coefficients are slightly different due to different responses to the variety
and temperature.

```{r moderately_resistant_model, echo=TRUE}
mr_model <- nema_long %>%
  filter(Variety == "GS50a") %>%
  quadratic_model()

par(mfrow = c(2, 2))
plot(mr_model)

summary(mr_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "GS50a") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x + poly(x, 2),
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("GS50a Quadratic Model")
```

# Discussion and Conclusions

As in the original paper, the model equations can be derived from these models
as well. The derived regression equations are:

Gatcher: $$ln(P. thornei + 1) = -0.000003(0.0000009)T^2 + 0.009(0.0019)T + 5.4671(0.904)$$

GS50a: $$ln(P. thornei + 1) = -0.000002(0.0000007)T^2 + 0.0063(0.0014)T + 5.1559(0.678)$$

Unplanted control: $$ln(P. thornei + 1) = 0.0013(0.00018)T + 5.4151(0.193)$$

# Reproducibility
```{r reproducibility, echo=FALSE}
devtools::session_info()
```
